Software Quality Assurance in INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures

### 2019-12-08 paper-jgc.tex
minor revision changes

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003)
main goal was to provide scientific communities (including physics, biology and
earth sciences) with intensive computing and large--scale dataset analysis capabilities.
The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
Grid was the emerging technology to be used in order to address their requirements,
thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly
based on Globus middleware components and services \cite{globus}. The project was organized in 12 work packages, from
which 5 were devoted to software development and coordinated by an
Architecture Task Force, supervising the overall design and technical consistency
of the developments.

@@

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they reduce the time in which a given defect is exposed or has an impact, allowing the
development teams to act promptly based on the regular feedback \cite{chen2015}. The software updates of
{\sl INDIGO--DataCloud} products, which have been taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories. Nonetheless,
expert supervision to validate the results is required.

@@

All along the project we have observed that source code review is the cornerstone to
improve the effectiveness of the unit and functional -- when applicable -- test cases. 
{\sl INDIGO--DataCloud} source code repositories are protected against direct pushes 
so that the code review stage always takes place. Code reviewers analyse the content
of each change in the source code and the suitability -- according to the SQA criteria -- of the
associated tests, being the last checkpoint before the change is added into the
production branch. Hence the vital importance of providing the tests alongside the
code for every given change.
% ?? figure with code reviews or sth about PRs, external participation
%Functional quality, on the other hand, was assessed within the {\sl INDIGO--DataCloud} 
%software lifecycle through the per--change execution of the functional tests provided 
%by the developers and the ultimate stage of integration testing in the aforementioned 
%preview testbed. As we have seen, the latter testing strategy was entirely performed 
%in a manual way, and appeared as a helpful tool to spot defects after exposing the 
%deployed software to interactions with external components. Additionally, the functional 
%tests, were primarily useful to check that the software met the design specifications, 
%and most importantly, that previously working functionalities are preserved when new 
%features are added. As mentioned, these type of testing is meant to be continuously 
%executed in an automated manner using the CI infrastructure. 
However, by the time of
{\sl INDIGO--1} only 30\% of the total of products provided automated functional tests,
reaching almost 60\% in {\sl INDIGO--2}. Although test reports were asked in those cases
where automated functional tests were not provided, the fulfillment of 
their fundamental functional requirements was tougher to track, resulting in a non--viable
solution for assessing minor changes, especially when analysing the regressions. The major issues observed that hindered developers from the 
implementation of automated functional tests were related to the unavailability
-- at the time -- of libraries and frameworks for tackling specific tests -- such as 
testing graphical interfaces --, or in some cases, the unawareness about their existence.
We infer from this experience that automated functional tests should be a goal in modern
software development as they complement the structural quality checks by adding the
validation of the software's identified functionalities, key in the operational impact 
and user acceptance. 
% figure with % of products with functional tests
% ?? products not developed from scratch challenged the application of SQA policies??

@@

\bibitem{indigo-d31}
J. Gomes \emph{et al.},
``Initial Plan for WP3,''
\emph{INDIGO--DataCloud Deliverable 3.1}.
[Online]. Available: \url{https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31}.
[Accessed 14 Feb. 2019].

>>>>

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003)
main goal was to provide scientific communities (including physics, biology and
earth sciences) with intensive computing and large--scale dataset analysis capabilities.
The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
Grid was the emerging technology to be used in order to address their requirements,
thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly
based on Globus middleware components and services \cite{globus}.

@@

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they reduce the time in which a given defect is exposed or has an impact, allowing the
development teams to act promptly based on the regular feedback \cite{chen2015}. A total of 12 software
updates --including 91 product releases \cite{indigo-d33}--, which have been taking place since {\sl INDIGO--1}, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories. Nonetheless,
expert supervision to validate the results is required.

@@

All along the project we have observed that source code review is the cornerstone to
improve the effectiveness of the unit and functional -- when applicable -- test cases. 
{\sl INDIGO--DataCloud} source code repositories are protected against direct pushes 
so that the code review stage always takes place. Code reviewers analyse the content
of each change in the source code and the suitability -- according to the SQA criteria -- of the
associated tests, being the last checkpoint before the change is added into the
production branch. Hence the vital importance of providing the tests alongside the
code for every given change.
% ?? figure with code reviews or sth about PRs, external participation
%Functional quality, on the other hand, was assessed within the {\sl INDIGO--DataCloud} 
%software lifecycle through the per--change execution of the functional tests provided 
%by the developers and the ultimate stage of integration testing in the aforementioned 
%preview testbed. As we have seen, the latter testing strategy was entirely performed 
%in a manual way, and appeared as a helpful tool to spot defects after exposing the 
%deployed software to interactions with external components. Additionally, the functional 
%tests, were primarily useful to check that the software met the design specifications, 
%and most importantly, that previously working functionalities are preserved when new 
%features are added. As mentioned, these type of testing is meant to be continuously 
%executed in an automated manner using the CI infrastructure. 
However, by the time of
{\sl INDIGO--1} only 30\% of the total of products provided automated functional tests,
reaching almost 60\% in {\sl INDIGO--2}. This fact was undoubtedly the main cause that challenged the full implementation of 
the described DevOps approach. 
Although testing reports were asked for those components that did not provide automated functional tests, the fulfillment of 
their fundamental functional requirements was tougher to track, resulting in a non--viable
solution for assessing minor changes, especially when analysing the regressions. The major issues observed that hindered developers from the 
implementation of automated functional tests were related to the unavailability
-- at the time -- of libraries and frameworks for tackling specific tests -- such as 
testing graphical interfaces --, or in some cases, the unawareness about their existence.
We infer from this experience that automated functional tests should be a goal in modern
software development as they complement the structural quality checks by adding the
validation of the software's identified functionalities, key in the operational impact 
and user acceptance. 
% figure with % of products with functional tests
% ?? products not developed from scratch challenged the application of SQA policies??

@@

\bibitem{indigo-d31}
J. Gomes \emph{et al.},
``Initial Plan for WP3,''
\emph{INDIGO--DataCloud Deliverable 3.1}.
[Online]. Available: \url{https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31}.
[Accessed 14 Feb. 2019].
\bibitem{indigo-d33}
J. Gomes \emph{et al.},
``WP3 Final Report,''
\emph{INDIGO--DataCloud Deliverable 3.3}.
[Online]. Available: \url{https://www.indigo-datacloud.eu/documents/wp3-final-report-d33}.
[Accessed 14 Feb. 2019].
